Bandwidth-Aware Reconfigurable Memory Design with Hybrid Memory Technologies

### 2011-04-09 trunk/paper.tex

The bandwidth problem of multicore processors has drawn much attention recently.
Yu \emph{et al.} proposed a LLC partitioning algorithm to minimize bandwidth
requirement to off-chip main memory~\cite{Yu:2010:OCMB}. While most cache
partitioning techniques focus on cache miss rates, our work takes a different
approach in which tasks¡¯ memory bandwidth requirements are taken into account
when identifying a cache partitioning for multi-programmed and/or multithreaded
workloads. Cache resources are allocated with the objective that the overall
system bandwidth requirement is minimized for the target workload. The key
insight is that cache miss-rate information may severely misrepresent the actual
bandwidth demand of the task, which ultimately determines the overall system
performance and power consumption. <the moguls paper, the figure 1, however design for
induvidual application, i.e., with different application can generate different
designs.>

A large body of recent research focuses on exploring new memory technologies to
trade off between latency, bandwidth, and cost. Wu et al. explored various
memory technologies - SRAM, eDRAM [23], magnetic RAM (MRAM) [12] and PCM to best
construct L3 caches in terms of performance and power [24]. The eDRAM and MRAM
technologies are evaluated as LLCs by stacking each memory on top of the
processor die. Both studies mainly focused on reducing the latency gap between L2
cache~(LLC) and external memory and do not examine the bandwidth. <nvm cache
research: most for latency, not bandwidth opt.>

@@

Memory reconfiguration relies on accurate predictions of demand bandwidth of a
workload in response to the dynamically varying application characteristics.
Conventional last value or table based predictors can model neither long range
patterns of an application nor patterns with variable lengths. In this work, we
employ a statistical predictor to support for the BAROM design. The basic idea
of the predictor is borrowed from the n-Gram models, which are typically used by
speech and natural language processing~\cite{smoothing-tech}. The n-gram models
are usually formulated as a probability distribution $p(s)$ over a strings $s$,
and attemp to reflect how frequently the string occurs as a sentense. Prediction
on a reasonable combination of strings is then generated based on the
probability distribution. Our predictor employs the same basic idea, but
simplifies the model for performance considerations. In our model, each demand
bandwidth (DBW) sample abtained in a time interval is analogous to a string in
language, and a given length (the ``order'') of samples is stored in a table as
a pattern. At each time interval, the prediction engine will update the pattern
table with the new DBW sample, and calculate probability of the updated pattern.
The probability of a pattern $s$ of length $l$ can be calculated without loss as
the product of a sequence of conditional probabilities using the equation shown
below.

@@

\begin{equation}
  \hat{p}(w_i|w_{i-n+1}^{i-1})=\frac{c(w_{i-2}^{i})}{c(w_{i-2}^{i-1})}\label{eq:pr3}
\end{equation}
where $c(w_{a}^{b})$ is the number of times that the sequence $w_a..w_b$ appears
in preceding samples.

The demand bandwidth in the next time interval is predicted based on the
probabilities.

Figure~\ref{fig:predictor} shows the hardware components of our prediction
engine, which includes a pattern look-up-table, a probability vector, and a set
of counters.

@@

hardware: global predictor sitting before caches
statistical prediction

hierarchical prediction

performance opt: lut long enough but not too long (num of entries, bits per
entry)

>>>>

The bandwidth problem of multicore processors has drawn much attention recently.
Yu \emph{et al.} proposed a LLC partitioning algorithm to minimize bandwidth
requirement to off-chip main memory~\cite{Yu:2010:OCMB}. While most cache
partitioning techniques focus on cache miss rates, their work takes a different
approach in which tasks¡¯ memory bandwidth requirements are taken into account
when identifying a cache partitioning for multi-programmed and/or multithreaded
workloads. Cache resources are allocated with the objective that the overall
system bandwidth requirement is minimized for the target workload. The key
insight is that cache miss-rate information may severely misrepresent the actual
bandwidth demand of the task, which ultimately determines the overall system
performance and power consumption. However, their work only focused on LLC (L2
cache). Our design targe is the overall on-chip memory system, which provides
more design dimensions and flexibility. An analytical performance model was
proposed by Sun et al. to explore the design space of memory hierarchies for
throughput computing [moguls]. However, their memory hierarchy design was
tailored for individual applications, which was undesirable for a real system
design.

A large body of recent research focuses on exploring new memory technologies to
trade off between latency, bandwidth, and cost. Wu et al. explored various
memory technologies - SRAM, eDRAM [23], magnetic RAM (MRAM) [12] and PCM to best
construct L3 caches in terms of performance and power [24]. The eDRAM and MRAM
technologies are evaluated as LLCs by stacking each memory on top of the
processor die. Both studies mainly focused on reducing the latency gap between
L2 cache~(LLC) and external memory, but do not examine the design from bandwidth
perspective.

@@

Memory reconfiguration relies on accurate predictions of demand bandwidth of a
workload in response to the dynamically varying application characteristics.
Conventional last value or table based predictors can model neither long range
patterns of an application nor patterns with variable lengths. In this work, we
employ a statistical predictor to support for the BAROM design. The basic idea
of the predictor is borrowed from the n-Gram models, which are typically used by
speech and natural language processing~\cite{smoothing-tech}. The n-gram models
are usually formulated as a probability distribution $p(s)$ over a strings $s$,
and attemp to reflect how frequently the string occurs as a sentense. Prediction
on a reasonable combination of strings is then generated based on the
probability distribution. Our predictor employs the same basic idea, but with
very different implementation considerations. While language modeling is built
from a set of previous collected training sentenses with finite lengths, our
statistical model needs to be able to dynamically generate prediction based on a
continuous sequence of metrics. In addition, we can only implement limited
resolution for the metrics. Therefore, we need to normalize the demand bandwidth
values with a limited number of quantization bins. In our model, each demand
bandwidth (DBW) sample abtained in a time interval is analogous to a string in
language, and a given length (the ``order'') of samples is stored in a table as
a pattern. At each time interval, the prediction engine will update the pattern
table with the new DBW sample, and calculate probability of the updated pattern.
The probability of a pattern $s$ of length $l$ can be calculated without loss as
the product of a sequence of conditional probabilities using the equation shown
below.

@@

\begin{equation}
  \hat{p}(w_i|w_{i-n+1}^{i-1})=\frac{c(w_{i-2}^{i})}{c(w_{i-2}^{i-1})}\label{eq:pr3}
\end{equation}
where $c(w_{a}^{b})$ is the number of times that the sequence $w_a..w_b$ appears
in preceding samples. The demand bandwidth in the next time interval is
predicted to be the last value in the pattern with the highest probability.

Figure~\ref{fig:predictor} shows the hardware components of our prediction
engine, which includes a pattern look-up-table, a probability vector, and a set
of counters. The storage overhead

@@

The computational complexity can be estimated as $O(ql)$, where $q$ is the
number of quantization bins.
